---
title: "Reproducible Data Analysis: Airports, Monte Carlo, and GenAI Planning"
author: "Sean Kennelly"
date: "June 12, 2024"

format:
  pdf:
    toc: true
    toc-depth: 2
    number-sections: true
    geometry: margin=1in
    fontsize: 11pt
---

```{R}
#| echo: false
#| include: false
# -----------------------
# AIRPORT DATA
# -----------------------
library(tidyverse)
library(rvest)
library(ggplot2)

url <- "https://en.wikipedia.org/wiki/List_of_busiest_airports_by_passenger_traffic"

# More robust table selection: find the table by its caption
page <- read_html(url)
table_captions <- page |> html_elements("caption") |> html_text()
busiest_table_index <- which(str_detect(table_captions, "(?i)busiest airports by passenger traffic"))[1]

# Check if we found a table, else use a fallback (e.g., tables[[1]])
if (is.na(busiest_table_index)) {
  warning("Target table not found by caption. Using first table as fallback.")
  airport_data <- html_table(page)[[1]]
} else {
  airport_data <- html_table(page)[[busiest_table_index]]
}

# Define airports of interest
airports_of_interest <- c(
  "Hartsfield–Jackson Atlanta International Airport", # Note: Using en-dash from Wikipedia
  "Frankfurt Airport",
  "Beijing Daxing International Airport",
  "Dubai International Airport",
  "Los Angeles International Airport",
  "Heathrow Airport"
)

# Clean and process data using parse_number for safety
airport_table <- airport_data |>
  filter(`Airport` %in% airports_of_interest) |>
  mutate(
    # readr::parse_number expertly handles commas, spaces, and text
    Passengers = parse_number(`Totalpassengers`)
  ) |>
  select(Airport, Country, Passengers) |>
  arrange(desc(Passengers)) # Sort the table for the report

# Prepare data for plotting
airport_plot_data <- airport_table |>
  mutate(Airport = fct_reorder(Airport, Passengers)) # Order bars by passenger count

# Create the plot object
airport_plot <- ggplot(airport_plot_data,
       aes(x = Airport, y = Passengers, fill = Country)) +
  geom_col() +
  coord_flip() + # Flip coordinates for better readability of long airport names
  labs(
    title = "Busiest Airports by Passenger Traffic",
    subtitle = "Comparison of selected global airports",
    x = NULL, # Remove x-axis label as it's obvious
    y = "Total Passengers (millions?)", # Suggest adding unit if known
    fill = "Country"
  ) +
  theme_minimal() +
  theme(legend.position = "bottom")

```

# Busiest Airports Analysis

## Summary Table

The table below shows six major international airports ranked by total passenger traffic. Hartsfield-Jackson Atlanta International Airport appears as the busiest airport among the selected group, handling the largest number of passengers.

Heathrow and Dubai International Airport also show very high traffic volumes, highlighting their importance as global travel hubs. European airports such as Heathrow and Frankfurt handle slightly fewer passengers but still play major roles in international travel.

Beijing Daxing International Airport appears lower in the ranking, which may be due to its relatively recent opening compared to the others.

```{r}
airport_table
```

## Visualization

The bar chart provides a visual comparison of passenger traffic across the six airports. Hartsfield-Jackson Atlanta stands out clearly with the highest passenger count.

The use of color to represent country helps highlight geographic distribution. U.S. airports dominate the list, while Europe and Asia are also represented by key international hubs.

Compared to the table, the plot makes it easier to quickly see differences in magnitude between airports.

```{r}
airport_plot
```

Alt Text: bar chart comparing passenger traffic at six major airports

Long Description: Bar chart comparing total passengers across six major airports. Dubai is busiest, followed by the UK and US. The chart uses vertical bars for easy comparison across airports and highlights the difference in passenger traffic.

------------------------------------------------------------------------

# Monte Carlo

```{R}
#| echo: false
# -----------------------
# MONTE CARLO SIMULATION
# -----------------------
monte_carlo_sim <- function(n, x_min, x_max, y_min, y_max, func) {
  
  # generate random x and y values within bounds
  x_vals <- runif(n, min = x_min, max = x_max)
  y_vals <- runif(n, min = y_min, max = y_max)
  
  # evaluate function at x values
  f_vals <- func(x_vals)
  
  # classify points
  position <- ifelse(y_vals <= f_vals, "below_or_on", "above")
  
  # return data frame
  data.frame(
    x = x_vals,
    y = y_vals,
    f_x = f_vals,
    position = position
  )
}

set.seed(42) # Set seed for reproducibility

mc_data <- monte_carlo_sim(
  n = 1000,
  x_min = 0,
  x_max = 20,
  y_min = 0,
  y_max = 0.16,
  func = function(x) dchisq(x, df = 3)
)

# Chi-square pdf function
chisq_pdf <- function(x, df = 3) {
  dchisq(x, df = df)
}

# wrangling and estimation
mc_results <- mc_data |>
  mutate(
    below_curve = ifelse(y <= f_x, 1, 0)
  )

# proportion under curve
p <- mean(mc_results$below_curve)

# bounds 
x_min <- min(mc_results$x)
x_max <- max(mc_results$x)
y_min <- min(mc_results$y)
y_max <- max(mc_results$y)

area_rect <- (x_max - x_min) * (y_max - y_min)

# Monte Carlo estimate of integral
integral_estimate <- p * area_rect

library(ggplot2)

mc_plot <- (ggplot(mc_results, aes(x = x, y = y)) +
  geom_point(aes(color = factor(below_curve)), alpha = 0.5) +
  stat_function(fun = chisq_pdf, color = "blue", linewidth = 1) +
  labs(
    title = "Monte Carlo Integration",
    subtitle = paste("Estimated Integral =", round(integral_estimate, 4)),
    x = "x",
    y = "y",
    color = "Below Curve"
  ) +
  theme_minimal()
  )
```

## Monte Carlo Numerical Integration

This section demonstrates how Monte Carlo simulation can be used to approximate the integral of a function. In this case, the function is the chi-square probability density function with 3 degrees of freedom.

Random points are generated within a rectangular region, and the proportion of points that fall below the curve is used to estimate the area under the curve.

## Results and Discussion

As the number of sampled points increases, the Monte Carlo estimate becomes more accurate. With a sample size of 1,000 points, the estimate is already reasonably close to the true value.

The visualization shows random points distributed above and below the curve. As sample size increases, the shape of the chi-square distribution becomes more apparent.

This method demonstrates the trade-off between accuracy and computation: larger samples give better estimates but require more computation.

Because randomness is involved, results may vary slightly, but reproducibility is ensured using a fixed seed (in code appendix).

```{r}
mc_plot
```

Alt Text: Monte Carlo chi-square integration with different sample sizes

Long Description: This figure shows four visualizations in a 2x2 format of Monte Carlo integration for the Chi-square distribution using sample sizes of 10, 100, 1000, and 10000. Each plot displays random points inside a rectangle and a blue chi-square curve. Points are colored based on whether they fall above or below the curve. As sample size increases, the points fill the curve and the estimate becomes more accurate.

------------------------------------------------------------------------

# Planning and Prompting GenAI

## Goals

Our goals for this question is to tidy and wrangle the calcium dataset to be able to produce visiualization.

## Needs

Nouns: calcium dataset, columns, rows, control group, treatment group Verbs: reshape, group, import, pivot, convert

## Step-by-Step

1.  Import calcium.csv
2.  Rename columns clearly if needed
3.  Add subject ID column
4.  Pivot longer so each row is: subject, group, time, calcium value
5.  Convert time to ordered factor
6.  display table

## GenAI Prompt

I used Microsoft copilot, and my prompt included the calcium.csv file and my goal, needs, and step-by-step plan.

## GenAI Response

Copilot:

Here’s a clean, step‑by‑step way to do this in R with tidyverse.

r library(tidyverse)

1.  Import calcium.csv

calcium_raw \<- read_csv("calcium.csv")

calcium_raw Your file starts like:

y0,y1,y2,y3,y0,y1,y2,y3 87.3,86.9,86.7,75.5,83.8,85.5,86.2,81.2

2.  Rename columns clearly Assume first 4 columns = control, last 4 = treatment.

r calcium \<- calcium_raw \|\> rename( y0_control = y0, y1_control = y1, y2_control = y2, y3_control = y3, y0_treat = y0_1, y1_treat = y1_1, y2_treat = y2_1, y3_treat = y3_1 ) If read_csv() auto-renamed duplicate names, they’ll be like y0, y1, y2, y3, y0_1, y1_1, etc.—adjust to match what you see in calcium_raw.

3.  Add subject ID column r calcium \<- calcium \|\> mutate(subject = row_number())

4.  Pivot longer to subject–group–time–value r calcium_long \<- calcium \|\> pivot_longer( cols = starts_with("y"), names_to = c("time", "group"), names_sep = "\_", values_to = "calcium" ) This uses names like y0_control → time = "y0", group = "control".

5.  Convert time to ordered factor r calcium_long \<- calcium_long \|\> mutate( time = factor(time, levels = c("y0", "y1", "y2", "y3"), ordered = TRUE), group = factor(group, levels = c("control", "treat")) )

6.  Display table r calcium_long \|\> arrange(subject, group, time) \|\> print(n = 20) You now have one row per subject–group–time–calcium, ready for visualization (e.g., ggplot(calcium_long, aes(time, calcium, color = group, group = interaction(subject, group))) + geom_line()).

## Compare and Contrast

The response followed the structure given in the prompt and stayed focused on the task. It clearly stated the goal and broke the work into steps. This is useful to having a better understanding of what the genai is telling us. It used pivot_longer() to reorganize the data. It also grouped by time, subject, and treatment group. This prepared the data for visualizations. A generic prompt would be less focussed. It would likely focus on basic cleaning and simple plotting. It would likely not consider how the dataset is structured. It would also miss the treatment vs control distinction. That would limit how useful the output is.

---

# Narrative

## Self-Reflection
Throughout this course, I have learned a lot about myself while working through these assignments. This course's emphasis on planning and presentation has helped me grow to see the whole process of our work and analysis. The goals, needs, and step-by-step process before writing our code for assignments really helps me write code easier and understand what to do. It breaks down the problem and makes the process of writing code especially for problems like the Monte Carlo problem much simpler. 

Understanding the importance of the presentation is also beneficial when completing projects. This assignment has helped me put together a lot of skills I have learned throughout the semester. For example, from creating a tidy table using R, to then displaying the plots and graphs on a pdf, this assignment challenged my skills. 

Formatting this pdf also helped me understand how important the layout of your work is before displaying. This helps keep our goals focused and our work understandable and reproducible. 

---

# Appendix A: GenAI Usage

## Usage 1: Copilot planning

## Tool:

Microsoft Copilot

## Date:

4/14/2026

## Prompt:

Goals

Our goals for this question is to tidy and wrangle the calcium dataset to be able to produce visiualization.

Needs

Nouns: calcium dataset, columns, rows, control group, treatment group Verbs: reshape, group, import, pivot, convert

Step-by-Step

1.  Import calcium.csv
2.  Rename columns clearly if needed
3.  Add subject ID column
4.  Pivot longer so each row is: subject, group, time, calcium value
5.  Convert time to ordered factor
6.  display table

calcium.csv

## Response: Copilot:

Here’s a clean, step‑by‑step way to do this in R with tidyverse.

r library(tidyverse)

1.  Import calcium.csv

calcium_raw \<- read_csv("calcium.csv")

calcium_raw Your file starts like:

y0,y1,y2,y3,y0,y1,y2,y3 87.3,86.9,86.7,75.5,83.8,85.5,86.2,81.2

2.  Rename columns clearly Assume first 4 columns = control, last 4 = treatment.

r calcium \<- calcium_raw \|\> rename( y0_control = y0, y1_control = y1, y2_control = y2, y3_control = y3, y0_treat = y0_1, y1_treat = y1_1, y2_treat = y2_1, y3_treat = y3_1 ) If read_csv() auto-renamed duplicate names, they’ll be like y0, y1, y2, y3, y0_1, y1_1, etc.—adjust to match what you see in calcium_raw.

3.  Add subject ID column r calcium \<- calcium \|\> mutate(subject = row_number())

4.  Pivot longer to subject–group–time–value r calcium_long \<- calcium \|\> pivot_longer( cols = starts_with("y"), names_to = c("time", "group"), names_sep = "\_", values_to = "calcium" ) This uses names like y0_control → time = "y0", group = "control".

5.  Convert time to ordered factor r calcium_long \<- calcium_long \|\> mutate( time = factor(time, levels = c("y0", "y1", "y2", "y3"), ordered = TRUE), group = factor(group, levels = c("control", "treat")) )

6.  Display table r calcium_long \|\> arrange(subject, group, time) \|\> print(n = 20) You now have one row per subject–group–time–calcium, ready for visualization (e.g., ggplot(calcium_long, aes(time, calcium, color = group, group = interaction(subject, group))) + geom_line()).

------------------------------------------------------------------------

# Appendix B: Code Appendix

## Airport Data Code
```{r}
#| eval: false
# -----------------------
# AIRPORT DATA
# -----------------------
library(tidyverse)
library(rvest)
library(ggplot2)

url <- "https://en.wikipedia.org/wiki/List_of_busiest_airports_by_passenger_traffic"

# More robust table selection: find the table by its caption
page <- read_html(url)
table_captions <- page |> html_elements("caption") |> html_text()
busiest_table_index <- which(str_detect(table_captions, "(?i)busiest airports by passenger traffic"))[1]

# Check if we found a table, else use a fallback (e.g., tables[[1]])
if (is.na(busiest_table_index)) {
  warning("Target table not found by caption. Using first table as fallback.")
  airport_data <- html_table(page)[[1]]
} else {
  airport_data <- html_table(page)[[busiest_table_index]]
}

# Define airports of interest
airports_of_interest <- c(
  "Hartsfield–Jackson Atlanta International Airport", # Note: Using en-dash from Wikipedia
  "Frankfurt Airport",
  "Beijing Daxing International Airport",
  "Dubai International Airport",
  "Los Angeles International Airport",
  "Heathrow Airport"
)

# Clean and process data using parse_number for safety
airport_table <- airport_data |>
  filter(`Airport` %in% airports_of_interest) |>
  mutate(
    # readr::parse_number expertly handles commas, spaces, and text
    Passengers = parse_number(`Totalpassengers`)
  ) |>
  select(Airport, Country, Passengers) |>
  arrange(desc(Passengers)) # Sort the table for the report

# Prepare data for plotting
airport_plot_data <- airport_table |>
  mutate(Airport = fct_reorder(Airport, Passengers)) # Order bars by passenger count

# Create the plot object
airport_plot <- ggplot(airport_plot_data,
       aes(x = Airport, y = Passengers, fill = Country)) +
  geom_col() +
  coord_flip() + # Flip coordinates for better readability of long airport names
  labs(
    title = "Busiest Airports by Passenger Traffic",
    subtitle = "Comparison of selected global airports",
    x = NULL, # Remove x-axis label as it's obvious
    y = "Total Passengers (millions?)", # Suggest adding unit if known
    fill = "Country"
  ) +
  theme_minimal() +
  theme(legend.position = "bottom")
```

## Monte Carlo Code
```{R}

#| eval: false
# -----------------------
# MONTE CARLO SIMULATION
# -----------------------
monte_carlo_sim <- function(n, x_min, x_max, y_min, y_max, func) {
  
  # generate random x and y values within bounds
  x_vals <- runif(n, min = x_min, max = x_max)
  y_vals <- runif(n, min = y_min, max = y_max)
  
  # evaluate function at x values
  f_vals <- func(x_vals)
  
  # classify points
  position <- ifelse(y_vals <= f_vals, "below_or_on", "above")
  
  # return data frame
  data.frame(
    x = x_vals,
    y = y_vals,
    f_x = f_vals,
    position = position
  )
}

set.seed(42) # Set seed for reproducibility

mc_data <- monte_carlo_sim(
  n = 1000,
  x_min = 0,
  x_max = 20,
  y_min = 0,
  y_max = 0.16,
  func = function(x) dchisq(x, df = 3)
)

# Chi-square pdf function
chisq_pdf <- function(x, df = 3) {
  dchisq(x, df = df)
}

# wrangling and estimation
mc_results <- mc_data |>
  mutate(
    below_curve = ifelse(y <= f_x, 1, 0)
  )

# proportion under curve
p <- mean(mc_results$below_curve)

# bounds 
x_min <- min(mc_results$x)
x_max <- max(mc_results$x)
y_min <- min(mc_results$y)
y_max <- max(mc_results$y)

area_rect <- (x_max - x_min) * (y_max - y_min)

# Monte Carlo estimate of integral
integral_estimate <- p * area_rect

library(ggplot2)

ggplot(mc_results, aes(x = x, y = y)) +
  geom_point(aes(color = factor(below_curve)), alpha = 0.5) +
  stat_function(fun = chisq_pdf, color = "blue", linewidth = 1) +
  labs(
    title = "Monte Carlo Integration",
    subtitle = paste("Estimated Integral =", round(integral_estimate, 4)),
    x = "x",
    y = "y",
    color = "Below Curve"
  ) +
  theme_minimal()
```
